Fast Fourier transformation of electromagnetic data for computationally expensive kernels

### 2021-02-22 ManuscriptTime.tex
Initiate Revision II
Point 8

\documentclass[extra, camera,%
    % onecolumn,   % comment for two columns
    % referee,     % uncomment for double lines
    final,       % uncomment for changes
]{gji}
\usepackage{timet}

@@

% Remove once done
% \newcommand{\itodo}[1]{\todo[color=blue!40!white, inline]{\sffamily #1}}
% \newcommand{\mtodo}[1]{\todo[inline]{\sffamily #1}}
% \newcommand{\etodo}[1]{\todo[color=blue!40!white]{#1}}
\newcommand{\dul}[1]{\underline{\underline{#1}}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\rmk}[1]{{\color{red}{#1}}}
\newcommand{\ohmm}{\ensuremath{\Omega\,}\text{m}\xspace}

@@

@@

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. The same applies for the reverse operation, the fewer times
required, the quicker we obtain frequency-domain results. Even though we limit
our analysis here to frequency-to-time transformations, most of the arguments
are reversible and applicable to time-to-frequency transformations. We decided
to use a regular spacing of frequencies on a log-scale, rather than an adaptive
scheme. This approach is favourable if a wide range of offsets is needed, as
the required frequencies change with offset and an adaptive frequency selection
is therefore often tailored to a single offset. Also, it allows for
straightforward parallelization over frequencies, which is not completely
possible with an adaptive scheme.

@@

There are three key-parameters that have to be defined: Minimum and maximum
frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies per decade
($n_\mr{dec}^f$). Trial-and-error with a 3D code is very time-intensive.
However, a simplified, layered model for the required survey setup and a fast
1D modeller makes it possible to estimate these parameters easily. To this end,
we created a graphical, interactive tool using \empymod, as shown in
Figure~\ref{fig:GUI-FFTLog-lin}.
%
\begin{figure}
  \centering
  \includegraphics[width=\cwidth]{01-GUI-FFTLog-lin}
  \caption{Interactive frequency selection for a user-provided layered model
    (the shown model parameters are described in the text). This example shows
    the impulse response at an offset of 5\,km, for which it uses FFTLog from
    0.001\,Hz to 10\,Hz with five frequencies per decade. Note that the error
    is clipped for values smaller than 0.01\,\% and bigger than 100\,\%.}
  \label{fig:GUI-FFTLog-lin}
\end{figure}
%

The shown example model is a marine scenario with 1\,km water depth of
resistivity $\rho=0.3\,\ohmm$ ($\rho = \sigma^{-1}$) and a 100\,m thick target
of 100\,\ohmm at 1\,km below the seafloor in a background of 1\,\ohmm. The
source is 50\,m above the seafloor, the receiver is on the seafloor, and the
response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the semi-analytical
responses. The blue circles indicate the actually computed responses and the
black dots the frequencies which are interpolated or set to zero. The resulting
time-domain response has a relative error of less than 1\,\% everywhere except
for very early times.

Figure \ref{fig:GUI-FFTLog-log} shows exactly the same on a logarithmic scale,
without the interactive GUI-related elements. It can be seen that with the
chosen frequencies the time-domain response starts to divert above about 100\,s
and below 0.3\,s. It also shows the oscillating high-frequency part, which is
hard to interpolate and we therefore set it to zero. Figure
\ref{fig:GUI-DLF-lin} is the same as Figure \ref{fig:GUI-FFTLog-lin}, but
transformed with the DLF method applying the 81-point sine-cosine filter from
\cite{GEO.09.Key}. The same frequencies were computed as in the FFTLog case and
the missing ones interpolated. The error of the corresponding time-domain
response is comparable, so either FFTLog or DLF can be used.
%
\begin{figure*}
  \centering
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02a-GUI-FFTLog-log}}%
  \label{fig:GUI-FFTLog-log}}\hfill
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02b-GUI-DLF-lin}}%
  \label{fig:GUI-DLF-lin}}\\
  \caption{(a) Same as Figure \ref{fig:GUI-FFTLog-lin}, but on a logarithmic
  scale. (b) Using DLF instead of FFTLog, but computing the same frequency
  range as for the FFTLog. The resulting time-domain response has comparable
  accuracy (compare the error to the error in Figure
  \ref{fig:GUI-FFTLog-lin}).}
  \label{fig:GUI-log-lin}
\end{figure*}
%
In the above example we used 20 frequencies, but that many would not be
required for the shown offset of 5\,km, a few of the lower and higher
frequencies could be left out. With this frequency-selection we can model a
wide range of offsets. This is shown in Figure \ref{fig:multi-offset}, where
the same kernels were used to yield the responses at offsets $r=1.5,3,6$, and
12\,km. We only need to compute 50\,\% of the required frequencies in the case
of the FFTLog, and only 14\,\% in the case of the DLF.
%
\begin{figure*}
  \centering
  \includegraphics[width=0.75\fwidth]{03-multi-offset}
  \caption{Normalized (a) frequency- and (b) time-domain responses for offsets
    $r=1.5,3,6$, and 12\,km using the parameters defined in Figure
    \ref{fig:GUI-FFTLog-lin}. The coloured symbols are the actually computed
    responses, the black dots are the responses which are set to zero (high
    frequencies) or interpolated (low frequencies). The grey curves are the
    analytical responses.}
  \label{fig:multi-offset}
\end{figure*}
%

The shortest offset defines the highest required frequency, and the largest
offset the lowest required frequency. So the more one can restrict the
necessary offset range, the fewer frequencies are needed. Another important
factor is how to interpolate and extrapolate from the computed frequencies to
the frequencies required for the Fourier transform. For the FFTLog only
extrapolation for higher and lower frequencies is required. The EM response
becomes highly oscillatory for high frequencies, which makes it very hard to
extrapolate the response to frequencies $f>f_\mathrm{max}$. However, if
$f_\mathrm{max}$ is chosen judiciously, the importance of higher frequencies
for the Fourier transform can be neglected and we can set those responses to
zero. The extrapolation of frequencies $f<f_\mathrm{min}$ can be changed to an
interpolation by assuming a zero imaginary response at zero frequency, and we
then use PCHIP to interpolate the missing frequencies (as we work on a
logarithmic scale we cannot choose 0\,Hz nor 0\,V/m, but instead take
$10^{-100}$\,Hz and $10^{-100}$\,V/m). In the case of the DLF method we also
have to interpolate in between the computed frequencies, for which we found it
better to use a cubic spline. As can be seen by comparing Figure
\ref{fig:GUI-FFTLog-lin} with Figure \ref{fig:GUI-DLF-lin}, using the FFTLog or
the DLF with the same actually computed frequencies results in very similar
responses.

@@

@@

Figure \ref{fig:fullspace} (a) shows the frequency-domain result, where the
blue dots are the computed responses and the black dots correspond to the
interpolated values or the values set to zero.
%
\begin{figure*}
  \centering
  \includegraphics[width=\fwidth]{04-fullspace}
  \caption{(a) Frequency- and (b) time-domain results for the homogeneous space
    model. The red lines are the analytical solutions, the blue circles are the
    actually computed responses, the black dots are the interpolated responses,
    and the dashed black line the obtained time-domain response. The errors are
    clipped for values smaller than 0.1\,\% and bigger than 100\,\%.}
  \label{fig:fullspace}
\end{figure*}
%
Most of the computed values stay below a relative error of 1\,\%, our chosen
adaptive gridding only starts to generate considerable errors at higher
frequencies. Figure \ref{fig:fullspace}(b) shows the corresponding time-domain
result, where the dashed black line is the result obtained by transformation of
the frequency-domain response, on top of the red line which is the analytical
result. The relative error is mostly below 1\,\%, except for early times.
However, for practical reasons that is more than enough.
Figure~\ref{fig:fullspace-log} shows the same on a logarithmic scale, with
times up to 10 seconds. It clearly shows that if later times are required, we
would need to adjust our Fourier transform parameters.
%
\begin{figure}
  \centering
  \includegraphics[width=\cwidth]{05-fullspace-log}
  \caption{Same as in Figure~\ref{fig:fullspace} (b), but on a logarithmic
    scale. To improve later times we would have to compute lower frequencies;
    to improve earlier times we would have to compute more frequencies per
    decade to get a better resolution.}
  \label{fig:fullspace-log}
\end{figure}
%
Note that for the gridding we chose $n_\delta=12$, which is very dense. This
was necessary because we are relatively close to the source. If the offsets of
interest are larger this factor can be lowered considerably; 3--4 is often
enough.

@@

The results are shown in Figure~\ref{fig:3d-result}: In (a) the 1D background
responses and the relative error using the semi-analytical result, and in (b)
the responses including the target.
%
\begin{figure*}
  \centering
  \includegraphics[width=\fwidth]{10-3d-result}
  \caption{Responses for the model outlined in Figure~\ref{fig:3d-model} using
    time-domain and frequency-domain computations, and for the layered
    background also the semi-analytical result. In the lower plot of (a) the
    relative error (\%) is shown in comparison to the semi-analytical result,
    and in (b) the normalized difference (\%) between the two 3D codes. The
    results of these plots are clipped for values smaller than 0.1\,\% and
    bigger than 10\,\%.}
  \label{fig:3d-result}
\end{figure*}
%
The background comparison shows that both 3D codes do an acceptable job with a
relative error of a few percents at most; the result obtained through
transformation seems to be better at early times. The reason is probably the
implemented backward Euler scheme in the time-domain code that has an error of
order one in time. We cannot compare the errors for the response that includes
the target for lack of an analytical solution. The 1D background model is only
included to show that there is a significant response from the target. We
therefore show the normalized difference (NRMSD) between the two responses
$R_1$ and $R_2$ as a percentage, where $\mr{NRMSD~(\%)} = 200|R_1 - R_2|/(|R_1|
+ |R_2|)$. The NRMSD between the two codes is below 1\,\% everywhere except for
early times. Both codes took roughly 4--5 minutes to compute the two models
(single thread). However, in this particular comparison, the main difference in
runtime is not frequency-domain computation vs. time-domain computation, but
iterative solver vs. direct solver.

>>>>

\documentclass[extra, camera,%
    onecolumn,   % comment for two columns
    referee,     % uncomment for double lines
    % final,       % uncomment for changes
]{gji}
\usepackage{timet}

@@

% Remove once done
% \newcommand{\itodo}[1]{\todo[color=blue!40!white, inline]{\sffamily #1}}
% \newcommand{\mtodo}[1]{\todo[inline]{\sffamily #1}}
% \newcommand{\etodo}[1]{\todo[color=blue!40!white]{#1}}
\newcommand{\dul}[1]{\underline{\underline{#1}}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\rmk}[1]{{\color{red}{#1}}}
\newcommand{\ohmm}{\ensuremath{\Omega\,}\text{m}\xspace}
\newcommand{\rtodo}{{\color{red}\texttt{TODO:}}\xspace}

@@

\rtodo \added{Add citation to R.R. (2021) regarding RA and speed-up; limitation
to switch-on/-off.}

\rtodo \added{Add notion about preference for f-domain for IP problems}

@@

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. \added{This applies primarily for expensive kernels such
as used in 3D modelling where the computation of the response of a single
frequency takes much longer than the transformation itself.} The same applies
for the reverse operation, the fewer times required, the quicker we obtain
frequency-domain results. Even though we limit our analysis here to
frequency-to-time transformations, most of the arguments are reversible and
applicable to time-to-frequency transformations. We decided to use a regular
spacing of frequencies on a log-scale, rather than an adaptive scheme. This
approach is favourable if a wide range of offsets is needed, as the required
frequencies change with offset and an adaptive frequency selection is therefore
often tailored to a single offset\added{ or a limited range of offset}. Also,
it allows for straightforward parallelization over frequencies, which is not
completely possible with an adaptive scheme.

@@

\rtodo \added{Start this section with the method; fmin, fmax, ndec; below we
interpolate with PCHIP, above we set to zero, in-between we interpolate with
cubic spline.}
\replaced{The actually computed frequencies in our approach are defined by
three key-parameters}{There are three key-parameters that have to be defined}:
\replaced{The m}{M}inimum and maximum frequency ($f_\mr{min}, f_\mr{max}$) and
the number of frequencies per decade ($n_\mr{dec}^f$). \added{This creates a
frequency range which is much smaller than the frequency range required for the
transform.} \replaced{A t}{T}rial-and-error \added{approach} with a 3D code is
very time-intensive. However, a simplified, layered model for the required
survey setup and a fast 1D modeller makes it possible to estimate these
parameters easily. \replaced{An example of this is shown in
Figure~\ref{fig:FFTLog-log}.}{To this end, we created a graphical, interactive
tool using \empymod, as shown in Figure~\ref{fig:FFTLog-log}.}
%
\begin{figure}
  \centering
  \includegraphics[width=\cwidth]{01-FFTLog-log}
  \caption{\replaced{Example}{Interactive} frequency selection for a
    \replaced{simple}{user-provided} layered model (the shown model parameters
    are described in the text). This example shows the impulse response at an
    offset of 5\,km, for which it uses FFTLog from 0.001\,Hz to 10\,Hz with
    five frequencies per decade. \deleted{Note that the error is clipped for
    values smaller than 0.01\,\% and bigger than 100\,\%.}}
  \label{fig:FFTLog-log}
\end{figure}
%
\replaced{It }{The shown example model} is a marine scenario with 1\,km water
depth of resistivity $\rho=0.3\,\ohmm$ ($\rho = \sigma^{-1}$) and a 100\,m
thick target of 100\,\ohmm at 1\,km below the seafloor in a background of
1\,\ohmm. The source is 50\,m above the seafloor, the receiver is on the
seafloor, and the response is the inline $x$-directed $E$-field. The left
subplot shows the imaginary part in the frequency domain and the right plot the
corresponding impulse response in the time domain. The red lines are the
semi-analytical responses. The blue circles indicate the actually computed
responses and the black dots the frequencies which are interpolated or set to
zero. The resulting time-domain response has a relative error of less than
1\,\% everywhere except for very early times.

Figure \ref{fig:FFTLog-lin} shows exactly the same on a
\replaced{linear}{logarithmic} scale\replaced{for the time and amplitude
axes}{, without the interactive GUI-related elements}. It can be seen that with
the chosen frequencies the time-domain response starts to divert above about
100\,s and below 0.3\,s. It also shows the oscillating high-frequency part,
which is hard to interpolate and we therefore set it to zero.
Figure~\ref{fig:DLF-lin} is the same as Figure \ref{fig:FFTLog-lin}, but
transformed with the DLF method applying the 81-point sine-cosine filter from
\cite{GEO.09.Key}. The same frequencies were computed as in the FFTLog case and
the missing ones interpolated. The error of the corresponding time-domain
response is comparable, so either FFTLog or DLF can be used.
%
\begin{figure*}
  \centering
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02a-FFTLog-lin}}%
  \label{fig:FFTLog-lin}}\hfill
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02b-DLF-lin}}%
  \label{fig:DLF-lin}}\\
  \caption{(a) Same as Figure \ref{fig:FFTLog-log}, but on a logarithmic
  scale. (b) Using DLF instead of FFTLog, but computing the same frequency
  range as for the FFTLog. The resulting time-domain response has comparable
  accuracy\deleted{ (compare the error to the error in Figure
    \ref{fig:FFTLog-log})}.}
  \label{fig:lin-lin}
\end{figure*}
%
In the above example we used 20 frequencies, but that many would not be
required for the shown offset of 5\,km, a few of the lower and higher
frequencies could be left out. With this frequency-selection we can model a
wide range of offsets. This is shown in Figure \ref{fig:multi-offset}, where
the same kernels were used to yield the responses at offsets $r=1.5,3,6$, and
12\,km. We only need to compute 50\,\% of the required frequencies in the case
of the FFTLog, and only 14\,\% in the case of the DLF.
%
\begin{figure*}
  \centering
  \includegraphics[width=0.75\fwidth]{03-multi-offset}
  \caption{Normalized (a) frequency- and (b) time-domain responses for offsets
    $r=1.5,3,6$, and 12\,km using the parameters defined in Figure
    \ref{fig:FFTLog-lin}. The coloured symbols are the actually computed
    responses, the black dots are the responses which are set to zero (high
    frequencies) or interpolated (low frequencies). The grey curves are the
    analytical responses.}
  \label{fig:multi-offset}
\end{figure*}
%

The shortest offset defines the highest required frequency, and the largest
offset the lowest required frequency. So the more one can restrict the
necessary offset range, the fewer frequencies are needed. Another important
factor is how to interpolate and extrapolate from the computed frequencies to
the frequencies required for the Fourier transform. For the FFTLog only
extrapolation for higher and lower frequencies is required. The EM response
becomes highly oscillatory for high frequencies, which makes it very hard to
extrapolate the response to frequencies $f>f_\mathrm{max}$. However, if
$f_\mathrm{max}$ is chosen judiciously, the importance of higher frequencies
for the Fourier transform can be neglected and we can set those responses to
zero. The extrapolation of frequencies $f<f_\mathrm{min}$ can be changed to an
interpolation by assuming a zero imaginary response at zero frequency, and we
then use PCHIP to interpolate the missing frequencies (as we work on a
logarithmic scale we cannot choose 0\,Hz nor 0\,V/m, but instead take
$10^{-100}$\,Hz and $10^{-100}$\,V/m). In the case of the DLF method we also
have to interpolate in between the computed frequencies, for which we found it
better to use a cubic spline. As can be seen by comparing Figure
\ref{fig:FFTLog-lin} with Figure \ref{fig:DLF-lin}, using the FFTLog or
the DLF with the same actually computed frequencies results in very similar
responses.

@@

\rtodo \added{Mention again that important for the transform are accurate
results; these gridding as what we implemented in the 3D code.}

@@

Figure \ref{fig:fullspace} (a) shows the frequency-domain result, where the
blue dots are the computed responses and the black dots correspond to the
interpolated values or the values set to zero.
%
\begin{figure*}
  \centering
  \includegraphics[width=\fwidth]{04-fullspace}
  \caption{(a) Frequency- and (b) time-domain results for the homogeneous space
    model. The red lines are the analytical solutions, the blue circles are the
    actually computed responses, the black dots are the interpolated responses,
    and the dashed black line the obtained time-domain response. \deleted{The
    errors are clipped for values smaller than 0.1\,\% and bigger than
    100\,\%.}}
  \label{fig:fullspace}
\end{figure*}
%
Most of the computed values stay below a relative error of 1\,\%, our chosen
adaptive gridding only starts to generate considerable errors at higher
frequencies. Figure \ref{fig:fullspace}(b) shows the corresponding time-domain
result, where the dashed black line is the result obtained by transformation of
the frequency-domain response, on top of the red line which is the analytical
result. The relative error is mostly below 1\,\%, except for early times.
However, for practical reasons that is more than enough.
Figure~\ref{fig:fullspace-log} shows the same on a logarithmic scale, with
times up to 10 seconds. It clearly shows that if later times are required, we
would need to adjust our Fourier transform parameters.
%
\begin{figure}
  \centering
  \includegraphics[width=\cwidth]{05-fullspace-log}
  \caption{Same as in Figure~\ref{fig:fullspace} (b), but on a logarithmic
    scale. To improve later times we would have to compute lower frequencies;
    to improve earlier times we would have to compute more frequencies per
    decade to get a better resolution.}
  \label{fig:fullspace-log}
\end{figure}
%
Note that for the gridding we chose $n_\delta=12$, which is very dense. This
was necessary because we are relatively close to the source. If the offsets of
interest are larger this factor can be lowered considerably; 3--4 is often
enough.

@@

The results are shown in Figure~\ref{fig:3d-result}: In (a) the 1D background
responses and the relative error using the semi-analytical result, and in (b)
the responses including the target.
%
\begin{figure*}
  \centering
  \includegraphics[width=\fwidth]{10-3d-result}
  \caption{Responses for the model outlined in Figure~\ref{fig:3d-model} using
    time-domain and frequency-domain computations, and for the layered
    background also the semi-analytical result. In the lower plot of (a) the
    relative error (\%) is shown in comparison to the semi-analytical result,
    and in (b) the normalized difference (\%) between the two 3D codes.
    \deleted{The results of these plots are clipped for values smaller than
    0.1\,\% and bigger than 10\,\%.}}
  \label{fig:3d-result}
\end{figure*}
%
The background comparison shows that both 3D codes do an acceptable job with a
relative error of a few percents at most; the result obtained through
transformation seems to be better at early times. The reason is probably the
implemented backward Euler scheme in the time-domain code that has an error of
order one in time. We cannot compare the errors for the response that includes
the target for lack of an analytical solution. The 1D background model is only
included to show that there is a significant response from the target. We
therefore show the normalized difference (NRMSD) between the two responses
$R_1$ and $R_2$ as a percentage, where $\mr{NRMSD~(\%)} = 200|R_1 - R_2|/(|R_1|
+ |R_2|)$. The NRMSD between the two codes is below 1\,\% everywhere except for
early times. Both codes took roughly 4--5 minutes to compute the two models
(single thread). However, in this particular comparison, the main difference in
runtime is not frequency-domain computation vs. time-domain computation, but
iterative solver vs. direct solver.

\subsection{\added{Induced Polarization}} % % % % % % % % % % % % % % % % % % %
